---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Either activity.zip or activity.csv file must exist in the working
directory to proceed
```{r}
if (!file.exists('activity.csv')) {unzip('activity.zip')}
activity = read.csv('activity.csv')
```



## What is mean total number of steps taken per day?
Let's explore the distribution of the number of steps taken each day. First, let's look at a histogram of the daily steps:

```{r}
daily_tot <- data.frame(tapply(activity$steps, activity$date, sum))
names(daily_tot) <- c("Steps")
hist(daily_tot$Steps, col='cornflowerblue', main='Distribution of daily step counts',breaks=10, xlab = 'Daily steps', ylab='Number of days')
```

Let's also look at the average and median number of steps per day, ignoring the days for which we have no data:
```{r}
mean(daily_tot$Steps, na.rm=TRUE)
median(daily_tot$Steps, na.rm=TRUE)
```



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
